A general and efficient representation of Ancestral Recombination Graphs

### 2023-10-06 paper.tex

\section*{ARGs and local trees}
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are said to be
``embedded'' within in ARG [CITES], and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of marginal distributions
(which could lead to interpretation as an integrated tree over many positions).

It is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% JK: this confused some people, so best left out?
% Technically, this structure is a ``forest'', as it allows us to have
% multiple disconnected trees, which occurs when all samples have not
% found a most recent common ancestor.
Oriented trees allow nodes
to have  any number of children; from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

The internal nodes of trees, however, are usually \emph{unlabelled},
because they most often correspond
to anonymous hypothetical ancestors. Internal
ARG nodes likewise correspond to anonymous hypothetical
ancestors, but if we do not label them we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Without labelled internal tree nodes,
we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.

\cite{shipilina2023origin} discuss the same ideas,
and note that
``full ARG ... contains more information than the series of tree
sequences along the genome''.
Their statement that an ARG
contains more information than its local trees is true only if
the local trees are unlabeled.
However, as we have shown, if the local trees
contain labelled internal (possibly unary) nodes,
then  there is precisely as
much information in them as an ARG.
Whether such unary nodes are \emph{observable}
is an entirely separate question, of course.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence of nodes that appear
``unary'' in one or more local trees
(where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions) and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
[FIXME this is clumsy; should mention somewhere, though.]
\citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are ``implicitly passing through points representing specific
individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

@@

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} arbitrary levels of precision
about recombination in a gARG.
It is also possible to \emph{infer} various levels of precision about
recombination.
To illustrate this, and to demonstrate the structural
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
The minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7 for this dataset~\citep{song2003parsimonious},
and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
The ARGs inferred by the different methods also have major differences
in how recombination information is represented.
Essentially,
\kwarg\ and \argweaver\ infer eARGs that contain
specific causal recombination \emph{events} for each breakpoint.
On the other hand, \tsinfer\ and
\relate\ do not infer causal events but rather
infer gARGs with differing levels of precision about recombination.
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see, for example, that the \relate\ trees ``share'' less topology
than other methods, but it certainly does not infer a series of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees; for example,
both infer that Af-f, Fr-f and Wa-f share a common ancestor along
the entire sequence.
Therefore, they contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importantance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2023extremely}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package
% FIXME-  more explicit?
(e.g.\ Relate, ARGneedle)
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.
% FIXME not sure where this goes
Earlier efforts to standardise ARG interchange
did not succeed~\citep{cardona2008extended,mcgill2013graphml}.

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
inference from data~\citep{kelleher2019inferring,wohns2022unified},
and calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank,nowbandegani2023extremely}.

@@

[TODO finish up with a positive note about interchange, and make the point
clearly that tskit is an ARG library. Splice in the stuff about haplotype
blocks and ARGON format here somewhere --- basically this is a derived product
of the gARG encoding, so while there are some computation advantages it's not
fundamental to the actual representation of the thing]
In summary, we advocate for decoupling between generative models that give rise to observed data and data formats that can faithfully and realistically encode these data. We have shown that ``genome'' ARG (gARG) encoding enables flexible encoding of today's datasets with realistic levels of precision. Importantly, genome ARG encoding should facilitate easier data interchange between different tools, which will in turn support active research in this area.

>>>>

\section*{ARGs and local trees}
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are
embedded within in ARG, and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of marginal distributions
(which could lead to interpretation as an integrated tree over many positions).

It is useful to have
a concrete definition and explicit process for local tree extraction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, then $\pi_u = -1$ (by convention).
% JK: this confused some people, so best left out?
% Technically, this structure is a ``forest'', as it allows us to have
% multiple disconnected trees, which occurs when all samples have not
% found a most recent common ancestor.
Oriented trees allow nodes
to have  any number of children; from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

The internal nodes of trees, however, are usually
considered to be \emph{unlabelled},
because they most often correspond
to anonymous hypothetical ancestors.
Internal ARG nodes likewise correspond to anonymous hypothetical
ancestors, but if we do not label them we lose the fact that
the \emph{same} ancestors persist across multiple trees.
Without labelled internal tree nodes,
we must infer their identity
from the tree topology, which is both
expensive to compute and imprecise. For example
we could infer that node \textsf{k} is shared in the second
and third trees of in Fig.~\ref{fig-simplification}D
because it subtends identical subtrees in both,
but it would not be straightforward to infer that it
is also the root of the first tree.

\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG ... contains more information than the series of tree
sequences along the genome''.
Their statement that an ARG
contains more information than its local trees is true only if
the local trees are unlabeled and/or do not contain unary nodes.
However, as we have shown, if the local trees
contain labelled internal (possibly unary) nodes,
then  there is precisely as
much information in them as an ARG.
Whether such unary nodes are \emph{observable}
is an entirely separate question, of course.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence of nodes that appear
``unary'' in one or more local trees
(where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals
% Following feedback from Wilder, this is more likely to confuse
% people unless we properly explain
% (or indeed cell divisions)
and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
% Not sure what to do with this - should mention though.
\citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

@@

The example of successive simplification of a simulated ARG
in Fig.~\ref{fig-simplification} illustrates the point
that is is possible to \emph{store} arbitrary levels of precision
about recombination in a gARG.
It is also possible to \emph{infer} various levels of precision about
recombination.
To illustrate this, and to demonstrate the structural
heterogeneity of the ARGs inferred  by current methods,
Fig.~\ref{fig-inferred-args} shows the results ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
for the classical \citet{kreitman1983nucleotide} benchmark dataset
using  \kwarg~\citep{ignatieva2021kwarg},
\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
\tsinfer~\citep{kelleher2019inferring},
and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
The minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7 for this dataset~\citep{song2003parsimonious},
and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
The ARGs inferred by the different methods also have major differences
in how recombination information is represented.
Essentially,
\kwarg\ and \argweaver\ infer eARGs that contain
specific causal recombination \emph{events} for each breakpoint.
On the other hand, \tsinfer\ and
\relate\ do not infer causal events but rather
infer gARGs with differing levels of precision about recombination.
The bottom row of Fig~\ref{fig-inferred-args} shows the extent along
the genome to which graph edges are shared between multiple trees.
We can see, for example, that the \relate\ trees ``share'' less topology
than other methods, but it certainly does not infer a series of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees; for example,
both infer that Af-f, Fr-f and Wa-f share a common ancestor along
the entire sequence.
They contain a level of detail about recombination
that lies somewhere between a sequence of unrelated local trees on one
extreme and a completely precise event ARG on the other.

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importantance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2023extremely}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
inference from data~\citep{kelleher2019inferring,wohns2022unified},
and calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
(Technically, edges in tree sequence terminology would be better
described as ``edge-intervals'', as each describes a single continguous
interval of genome inheritance between a pair of nodes. This
denormalisation of the gARG data model is for efficiency purposes.)
The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank,nowbandegani2023extremely}.

@@

[TODO finish up with a positive note about interchange, and make the point
clearly that tskit is an ARG library. Splice in the stuff about haplotype
blocks and ARGON format here somewhere --- basically this is a derived product
of the gARG encoding, so while there are some computation advantages it's not
fundamental to the actual representation of the thing]

Earlier efforts to standardise ARG interchange
did not succeed~\citep{cardona2008extended,mcgill2013graphml}.

In summary, we advocate for decoupling between generative models that give rise to observed data and data formats that can faithfully and realistically encode these data. We have shown that ``genome'' ARG (gARG) encoding enables flexible encoding of today's datasets with realistic levels of precision. Importantly, genome ARG encoding should facilitate easier data interchange between different tools, which will in turn support active research in this area.
